Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest

### 2019-05-20 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Add beta regression model for Fig 7
This should be final figure needed for monoculture flammability trials.

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel
  flammability. But less is known about the mechanisms by which litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Natural forests are often comprised of mixed stands,
  leading to mixtures of leaf litter on the surface fuelling the fires. Multiple
  studies have established the existence of non-additive effects in the
  flammability of dry litter, but we lack a mechanistic understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter. The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption and desorption. Burn trials were performed in moist litter beds of
  single species and of mixtures, to establish flammability response to the
  dry-down process. Moisture absorption is determined by traits pertaining to
  leaf geometry while moisture desorption is influenced by emergent properties
  of the litterbed. Species that produce more aerated litter beds (such as \emph{Q.
  kelloggii} and \emph{Pinus} species) have faster rates of litter desiccation. My results
  show a relationship between time since wetting and time to ignition and
  spread rate, with significant interaction with genus. This indicates a
  genus-specific behavior of flammability and time since wetting.
  We found non-additive effects in litter mixtures for both moisture content and
  flammability. Mixtures containing \emph{Q. kelloggii} and \emph{P. jeffreyi}, species that
  dry faster, produced more flammable behavior (lower time to ignition, faster
  spread rate, taller flames). Fire behavior of mixtures depended on moisture
  content, although mixtures without Q. kelloggii had always higher moisture
  content than predicted by the average fire behavior of the species in the
  mixture. At lower moisture contents (<75\%) the mixtures exhibit positive
  non-additive behavior and at higher moisture contents (> 200\%) they exhibit
  negative non-additive behavior.}

@@

During flammability trials, we recorded time to ignition (s), duration of
flaming combustion (s), spread rate (mm/s), calculated by dividing the length
of the burned surface by the time it takes the flaming front to reach the end
of the apparatus, maximal flame height (mm), and mass loss (\%), calculated as
the difference between fuel mass before and after flame extinction. Our
thermocouple system failed or data was lost for two days of burning trials,
therefore we have omitted those data from this analysis. Our past work
\cite{Magalhaes+Schwilk-2012} demonstrated that flame spread rate captures one
important axis of flammability variation \citep{Schwilk-2015,
  Pausas+Keeley+etal-2017} and percent fuel consumed captures a portion of the
second major axis of flammability variation (although not as well as
integrating temperature over time does). Our analysis, therefore, focused on
these two relatively orthogonal measures of flammability.

@@

As a consequnece of the relationships shown above, time since wetting had a strong effect on spread rate (p < 0.0001) and the slope of this effect varied by taxonimc (Fig. \ref{fig:spreadrate}, Table \ref{tab:spreadrate-anova}). The pines were able to ignite and burn much sooner after wetting than the other genera, none of which had measurable spread rates until burning was attempted 72 hours after wetting. At that point, the \emph{Abies}, \emph{Calocedrus}, and \emph{Sequiadendron} had much drying litter than did the \emph{Quercus}, yet the spread rates were slow.  \emph{Quercus} had the steepest response to time since wetting (Fig. \ref{fig5:spreadrate}).

\begin{table}[H]
  \caption[Mixed linear model of flame spread rate.]{Mixed model results for flame spread rate  as a function of time since wetting with genus as a fixed effect and species as a nested random effect. Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
  \label{tab:spreadrate-anova}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\input{figs_tables/spreadrate-anova.ltx}
\end{table}

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel
  flammability. But less is known about the mechanisms by which litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Natural forests are often comprised of mixed stands,
  leading to mixtures of leaf litter on the surface fuelling the fires.
  Multiple studies have established the existence of non-additive effects in
  the flammability of dry litter, but we lack a mechanistic understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter. The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption and desorption. Burn trials were performed in moist litter beds of
  single species and of mixtures, to establish flammability response to the
  dry-down process. Moisture dynamics vary across species with higher specific
  leaf area being associated with higher maximum absorption and lower litter
  bulk denisty being associated with faster drying rates. Species that produce
  more aerated litter beds (such as \emph{Q. kelloggii} and \emph{Pinus}
  species) have faster rates of litter desiccation. These differing moisture
  dynamics result in time since wetting having strongly differing effects on
  flammability across the major litter types.}

  % show a
  % relationship between time since wetting and time to ignition and spread rate,
  % with significant interaction with genus. This indicates a genus-specific
  % behavior of flammability and time since wetting. We found non-additive
  % effects in litter mixtures for both moisture content and flammability.
  % Mixtures containing \emph{Q. kelloggii} and \emph{P. jeffreyi}, species that
  % dry faster, produced more flammable behavior (lower time to ignition, faster
  % spread rate, taller flames). Fire behavior of mixtures depended on moisture
  % content, although mixtures without Q. kelloggii had always higher moisture
  % content than predicted by the average fire behavior of the species in the
  % mixture. At lower moisture contents (<75\%) the mixtures exhibit positive
  % non-additive behavior and at higher moisture contents (> 200\%) they exhibit
  % negative non-additive behavior.}

@@

During flammability trials, we recorded time to ignition (s), duration of
flaming combustion (s), flame spread rate (mm/s), calculated by dividing the
length of the burned surface by the time it takes the flaming front to reach
the end of the apparatus, maximal flame height (mm), and mass loss (\%),
calculated as the difference between fuel mass before and after flame
extinction. Some trials resulted only in smoldering combustion; for those
trials, a time to ignition and mass loss were recorded, but flame spread rate
and flame height were recorded as zero. Our thermocouple system failed or data
was lost for two days of burning trials, therefore we have omitted those data
from this analysis. Our past work \cite{Magalhaes+Schwilk-2012} demonstrated
that flame spread rate captures one important axis of flammability variation
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} and percent fuel consumed
captures a portion of the second major axis of flammability variation (although
not as well as integrating temperature over time does). Our analysis,
therefore, focused on these two relatively orthogonal measures of flammability.

@@

As a consequnece of the relationships shown above, time since wetting had a strong effect on spread rate (p < 0.0001) and the slope of this effect varied by taxonimc (Fig. \ref{fig:spread_time}, Table \ref{tab:spreadrate-anova}). The pines were able to ignite and burn much sooner after wetting than the other genera, none of which had measurable spread rates until burning was attempted 72 hours after wetting. At that point, the \emph{Abies}, \emph{Calocedrus}, and \emph{Sequiadendron} had much drying litter than did the \emph{Quercus}, yet the spread rates were slow.  \emph{Quercus} had the steepest response to time since wetting (Fig. \ref{fig:spread_time}).

\begin{figure}[H]
  \centering
\includegraphics[width=8cm]{figs_tables/fig6_spread_time.pdf}
\caption{Flame spread rate by time since wetting for litter of eight species. Model ANOVA table in Table \ref{tab:tab5_spread_time_anova} and coefficients in SI Table 4.
  \label{fig:spread_time}
\end{figure}

\begin{table}[H]
  \caption[Mixed linear model of flame spread rate.]{Mixed model results for flame spread rate  as a function of time since wetting with genus as a fixed effect and species as a nested random effect. Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
  \label{tab:spread_time_anova}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\input{figs_tables/tab5_spread_time_anova.ltx}
\end{table}
